---
title: "Day 6  Data Wrangling Tidyverse"
author: "Dalila Lara"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(skimr)
```

## Read in the World Bank data files 
```{r}
#wrangling data so missing values as ... are turned into na values
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), na = c("..", "")) 
wb_metadata <- read_csv(here("data", "wb_indicators_metadata.csv"))

#view data 
head(wb_indicators)
skim(wb_indicators)
```

## Tidy up my data 
pivot_longer + clean_names + separate()

Pivot longer to get years into a single columns
```{r}
#names(wb_indicators)
wb_indicators_long <- wb_indicators |>
  pivot_longer(cols = `2001 [YR2001]`:`2020 [YR2020]`,
               names_to = "year",
               values_to = "indicator_values")

#gives character not integer class
class(wb_indicators_long$year)

```

Separate to clean up year column 
  - Select columns
  - Filter rows
```{r}
# Separating year column 
# Using select function to further remove unwanted columns
# Class of year column is still character 
#head(wb_clean)
# Now we need to convert to integers with mutate as seen  (53)

wb_clean <- wb_indicators_long |>
  separate(col = year,
           into = c("year", "year_chr"), sep = " ") |>
  select(-year_chr, -`Series Code`, -`Country Code`) |>
  mutate(year = as.numeric(year)) |> # or create new column year_new 
  drop_na(`Series Name`) |> #drops na data from specified column 
  #Widen the variables currently trapped in `Series Name` to be spread across multiple columns, and get the values that populated the columns from `indicator_value`
  pivot_wider(names_from = `Series Name`, values_from = `indicator_values` ) 


```


```{r}
#List all column possible values of variable 
unique(wb_clean$`Series Name`)

```

```{r}
#renaming column names 
names(wb_clean) <- c("country", "year", "access_clean_fuels_pp", "access_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_stress")

```

Now making subset of 4 countries 
```{r}
#filter function
wb_subset <- wb_clean |>
  filter(country %in% c("Algeria", "Barbados", "Bulgaria", "Chile"))

#chcks if filter function was successful
unique(wb_subset$country)
```


Make a ggplot 
```{r}
ggplot(data = wb_subset, aes(x = year, y = co2_emissions_kt)) +
  geom_line(aes(color = country)) +
  facet_wrap(~country)
```



# Lecture 12 

## Filter() rows examples

### Examples: Filter to only keep exact matches (single matches)
`filter()` to extract specified rows 
```{r}
# Example 1 
# height is greater than 180
ex_1 <- starwars |>
  filter(height > 180)

#check minimum of subset 
min(ex_1$height)
```

```{r}
# Example 2
# eye color is equal to character blue
ex_2 <- starwars |>
  filter(eye_color == "blue")

#check unique()
unique(ex_2$eye_color)
```

```{r}
# Example 3

#only contains homeworld Naboo
ex_3 <- starwars |>
  filter(homeworld == "Naboo")

#check
unique(ex_3$homeworld)
```

### Examples (multiple matches )
Filtering to satisft multiple conditions
```{r}
# Example 4
#height is > 180 and the homeworld is Tatooine 
ex_4 <- starwars |>
  filter(height > 180 & homeworld == "Tatooine")
```

```{r}
# Example 5 
# Create a subset where hair color is brown and species is human 
ex_5 <- starwars |>
  filter(hair_color == "brown" & species == "Human")
```

### Examples if any statements are true
Filtering or statements `or`
```{r}
# Example 6
# height > 180 or eye color is yellow
ex_6 <- starwars |>
  filter(height > 180 | eye_color == "yellow")
```

```{r}
# Example 7
# Subset where rows are kept if homeworld is Endor or species is droid
ex_7 <- starwars |>
  filter(homeworld == "Endor" | species == "Droid")
```


### Writing an or statement in reference to a single variable
```{r}
# Example 8
# eye color blue, red, brown
ex_8 <- starwars |>
  filter(eye_color == "blue" | eye_color == "red" | eye_color == "brown")

# or exact same thing simplified `%in`
#Example 9
ex_9 <- starwars |>
  filter(eye_color %in% c("blue", "brown", "red"))
```

### Never do
```{r}
# Example 10
#asking if anything matches row in the given order of vector 
ex_10 <- starwars |>
  filter(eye_color == c("blue", "brown", "red"))
```

```{r}
#Example 11
# homeworld is Endor OR Tatoooine OR Naboo and height is < 150
ex_11 <- starwars |>
  filter(homeworld == "Endor" | homeworld == "Tatooine" | homeworld == "Naboo" & height < 150)
```

### Examples with the negate ! 

```{r}
#Example 12 
#Species does not equal human
ex_12 <- starwars |>
  filter(species != "Human")

#Example 13
ex_13 <- starwars |>
  filter(!species %in% c("Human", "Ewok", "Wookie"))
```

## `select()` select or exclude columns
```{r}
# Example 14 
ex_14 <- wb_clean |>
  select(country, year, co2_emission_kit)

#Example 15
ex_15 <- wb_clean |>
  select(year:fossil_fuel_cons_pct)

#Example 16
ex_16 <- wb_clean |>
  select(year:fossil_fuel_cons_pct, - access_electricity_pp)
```

```{r}
#Example 17
ex_17 <- wb_clean |>
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |>
  select(country:fossil_fuel_cons_pct)

#check
unique(ex_17$country)
names(ex_17)
```

```{r}
#keeps 3 cols. and renames these 3 
ex_18 <- wb_clean |>
  select(year_new = year,
        emissions = co2_emissions_kt,
        country_name = country)

```

```{r}
#example 18
#wb_clean selecto only colomuns for country, tears, and water stress and reorder as year, country, waterstress and filter to only include observations for any three countries

# Example 18
ex_18 <- wb_clean |> 
  select( year, country, water_stress) |> 
  filter(country %in% c("Mexico", "Chile", "Albania"))

#check
unique(ex_18$country)
```


## rename 
```{r}
#Example 20
#only these two columns get renamed 
ex_20 <- wb_clean |>
  rename(ffcp = fossil_fuel_cons_pct,
         ws = water_stress)

#rename any two columns
ex_21 <- wb_clean |>
  rename(ae_pp = access_electricity_pp,
         co2_e_kt = co2_emissions_kt)

```

## `mutate` update an exsisting column, or add a new one 
if you assign a NEW column name, it will add a new column to the end of your data frame. If you assign an exsisting column name, it will overwrite that column. 
```{r}
#adding a new column to wb_clean (tons)
#Example 22
ex_22 <- wb_clean |>
  mutate(co2_emissions_t = co2_emissions_kt * 1000) |>
  relocate(co2_emissions_t, .after = co2_emissions_kt)


#Example 23
#New column that just contains yay
ex_23 <- wb_clean |> 
  mutate(yay = "YAY")

#example 24 
ex_24 <- wb_clean |> 
  mutate(year = as.character(year))

class(ex_21$year)

#Ex 25
ex_25 <- ex_24 |>
  mutate(year = as.numeric(year))
```

## dplyr::relocate
```{r}
#changing order of columns
ex_26 <- wb_clean |> 
  relocate(water_stress, .after = year)
```

## `group_by()` |> `summarize` functions 

`group_by()` Creates groups under the hood based on variables you specify
`summarize()` apply your function to EACH GROUP and return them in a nice table 
```{r}
# Summarize data of specified group in table
ex_27 <- starwars |>
  filter(homeworld %in% c("Naboo", "Tatooine")) |>
  group_by(species, homeworld) |>
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))
```

```{r}
#Example 28
#Make a data frame
df <- tribble(
  ~species, ~length, ~count,
  "lobster", 12, 4,
  "lobster", 16, 6,
  "lobster", 15, 2
)

#make tidy by giving each observation its own row
df_case <- df |>
  uncount(count)
```

```{r}
# count function
#counts observations combo of group_by and summarize n()
starwars |>
  count(homeworld, species)


#gives same output as above 
starwars |>
  group_by(homeworld, species) |>
  summarize(size = n())
```

Practice Example 
- start from wb_clean
- keep observations for Nicaragua, Ecuador, Peru, Chile
- and for years > 2005
- keep column for country, year, and co2 emissions
- add new column that is co2 emissions converted to megatons
- find total co2 emissions still in data set by country

```{r}
practice_example <- wb_clean |>
  filter(country %in% c("Nicaragua", "Ecuador", "Peru", "Chile") &
           year > 2005) |>
  select( year, country, co2_emissions_kt) |> 
  mutate(co2_emissions_mt = co2_emissions_kt / 1000) |> 
  group_by(country) |>
  summarize(total_co2_mt = sum(co2_emissions_mt, na.rm = TRUE))

#check year > 2005
min(practice_example$year)
names(practice_example)
```

